A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-24 paper.tex

Prospective ARGs occur in individual-based forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree},
where the genetic inheritance between genomes is recorded
exhaustively and periodically
``simplified'' to retain only
the ancestry relevant to the current population.
\citet{kelleher2018efficient} refer to the structure
recording all such forwards-in-time genetic relationships
as an ``embellished pedigree`` (and suggest the term ``nedigree'').
Describing this instead as a ``prospective ARG'', and emphasising
the use of the same data structures and direct connection with
the usual retrospective interpretation of an ARG, may be
more useful.
The ``simplify'' algorithm which converts a prospective
ARG into a retrospective one is a powerful tool,
with other applications as we discuss in the following sections.

\section*{Ancestry resolution}\label{Ancestry_resolution}

These genomic intervals, directly ancestral to the samples
and  scattered across a set of lineages at some time in the past,
are usually known as the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

% TODO majorly condense this caption, it's much too long.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
two gARGs with alternative edge annotation schemes.
% (A) A conventional Griffiths eARG. Square nodes represent events, with
% each recombination (red) annotated with the position of a single breakpoint
% along the 7-unit-long chromosome.
(A) A gARG with edges annotated by . Oval nodes represent
genomes and annotations reflect ``local'' transmission of all genetic material
between parent and child. Edges that do not span the full genome are visually
highlighted, showing how breakpoint positions can be encoded.
(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
algorithm has been used to shorten edge spans to
reflect only the upwards transmission of genetic material from the samples
\textsf{a}, \textsf{b}, and \textsf{c}; very few edges then span the entire genome.
The depiction of the genome within each
node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
the relative importance of different transmission pathways to the samples. Nodes
\textsf{g},  \textsf{j}, \textsf{o} and \textsf{q} are redundant in that they correspond
to events which have no impact on the genetic ancestry of the samples.
}
\end{figure}

ARGs are most often thought of as a \emph{retrospective} structure however,
defined with respect to some set of (usually, contemporary) sample genomes.
In this case we are not explicitly interested in the
entire interval of genome descendant $u$ inherited from ancestor $v$,
but rather the subset of that interval that is \emph{ancestral to our
sample}.
Fig~\ref{fig-ancestry-resolution}B shows the graph that we obtain when
reduce the annotation intervals to those that are ancestral to the
sample (note that node $q$ is now ommitted because the MRCA of the sample
has already been found for the full genome at node $p$).
Tracking such
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
is much more useful~\citep{mcgill2013graphml,arenas2013importance}
and, indeed, all that is possible to
have information about when inferring ARGs from contemporary samples.
We refer to a gARG with inheritance annotations of this type as
``sample resolved''.

% Keeping track of resolved ancestral material is one of the distinguishing
% features of ARGs vs phylogenetic
% networks~\citep{mcgill2013graphml,arenas2013importance}, but
% is actually incompatible with the Griffiths gARG encoding (it does this
% implicitly, I guess)

@@

Given a gARG with direct-inheritance edge annotations
(Fig~\ref{fig-ancestry-resolution}B),
 we obtain the sample-resolved edge annotation
(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
Proceeding in topological order (such that an ancestor node is visited
after all of its descendants) the sample-resolved material for a node is the
union of the sample-resolved intervals on its inbound edges. We then annotate
the outbound edges with the intersection of their direct-inheritance intervals
and the just-computed sample-resolved intervals for the node.
For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
joining \textsf{i} to its ancestor \textsf{k} then carries
material on $(0, 2]$ since although \textsf{i} directly inherits interval
$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
ancestral to the sample. For each node, we also keep track of the number of
descendant samples over each genomic interval, and stop tracking that
interval once the lineages from all samples have coalesced. For example,
the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
already fully coalesced (indicated by a fully black genomic region).

@@

The process of converting a gARG with local-inheritance annotations
to its sample-resolved equivalent is also known as ``simplification'',
and is the key algorithm underlying recent
advances in forwards-in-time
simulation~\citep{kelleher2018efficient,haller2018tree}.
It has many useful properties, as we discuss in the next section.

>>>>

Prospective ARGs occur in individual-based forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree},
where the genetic inheritance between genomes is recorded
exhaustively and periodically
``simplified'' to retain only
the ancestry relevant to the current population.
\citet{kelleher2018efficient} refer to the structure
recording all such forwards-in-time genetic relationships
as an ``embellished pedigree`` (and suggest the term ``nedigree'').
Describing this instead as a ``prospective ARG'', and emphasising
the use of the same data structures and direct connection with
the usual retrospective interpretation of an ARG, may be
more useful.
The ``simplify'' algorithm which converts a prospective
gARG into a retrospective one is a powerful tool,
with several other applications
as we discuss in the following sections.

\section*{Converting an eARG to a gARG}
The process of converting an eARG to a gARG is very similar to
simplifying a prospective gARG with respect to a set of samples,
and highlights some important differences between the encodings.
The process involves two steps, and is illustrated in
Fig~\ref{fig-ancestry-resolution}.

\begin{figure}
\centering
% TODO update the source of this figure to make less tall. It's taking
% up far too much vspace at in native coords at the moment.
\includegraphics[width=0.75\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
Converting the \citet[][Fig.~1]{wiuf1999recombination} example graph
to a gARG. (A) The original eARG; square nodes represent events, with
each recombination (red) showing the corresponding breakpoint.
(B) The corresponding gARG with breakpoints directly converted to
edges annotated with inheritance intervals.
(C) The sample-resolved gARG resulting from simplifying with respect
to the samples, $a$, $b$, and $c$. Dashed lines show edges that are
no longer present (in practice, nodes $g$ and $q$ would also be removed).
Coalescence with respect to the sample is indicated by shaded bars, as
in Fig~\ref{fig-arg-in-pedigree}A.
Line thickness is proportional to the genomic span of each edge.
}
\end{figure}

The first step is straightforward
(Figure~\ref{fig-ancestry-resolution}B).
We duplicate the nodes and
edges from the input eARG, and the add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $(0,L]$.
If the node is a recombination node with a breakpoint $x$,
we annotate the outbound edges with the intervals $(0, x]$
and $(x, L]$.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

Ancestral material is a critical concept in a retrospective ARG,
and refers to the genomic intervals directly ancestral to the samples
scattered across lineages (the edges of the graph)
at some time in the past. At recombination
events, ancestral material is split between the ancestral lineages.
As genome segments coalesce in a common
ancestor the total amount ancestral material is reduced, until
eventually as we traverse pastwards through the graph, all
samples reach a most recent common ancestor at all locations
along the genome. The process of transmitting ancestral material
along the edges of an ARG is directly analogous to Hudson's
simulation algorithm for the coalescent with
recombination~\citep{hudson1983testing,kelleher2016efficient}.

Fig~\ref{fig-ancestry-resolution}C shows the graph that we obtain
by resolving the ancestral material transmitted along each edge
(using the ``simplify'' algorithm), and highlights some important
differences between the eARG and gARG encodings.
Firstly, we can see that some nodes and edges are removed entirely
from the graph. The ``grand MRCA'' $q$ is ommitted from the
fully resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between $g$ and $j$ is omitted because the recombination
event at position $5$ fell in non-ancestral material.
(This example of an eARG
from \citet{wiuf1999recombination} is something of a straw-man,
in that it is highly unlikely that an inference or simulation
program would produce an ARG that contains nodes or edges
that are not ancestral to the sample.)
More generally, the gARG encoding allows for ``local'' inspection
(with respect to a set of samples)
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the effects
of multiple events can be reasoned about at once. Many algorithms
that we wish to perform on an ARG will require resolving
the genetic material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to do this once and to store the result, whereas
the eARG encoding requires us to repeat the process.

%(see Appendix XXX) and has several features that we are unlikely to simulate or
%infer in practice. For example, by definition the samples contain no genetic information about
%nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
%Likewise, is is unlikely we would have information to infer the ``grand MRCA''
%\textsf{q} as all regions of the genome have already reached a
%common ancestor separately (in nodes \textsf{k} and \textsf{p}).

% reduce the annotation intervals to those that are ancestral to the
% sample (note that node $q$ is now ommitted because the MRCA of the sample
% has already been found for the full genome at node $p$).
% Tracking such
% ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
% is much more useful~\citep{mcgill2013graphml,arenas2013importance}
% and, indeed, all that is possible to
% have information about when inferring ARGs from contemporary samples.
% We refer to a gARG with inheritance annotations of this type as
% ``sample resolved''.

@@

% Given a gARG with direct-inheritance edge annotations
% (Fig~\ref{fig-ancestry-resolution}B),
%  we obtain the sample-resolved edge annotation
% (Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
% from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
% Proceeding in topological order (such that an ancestor node is visited
% after all of its descendants) the sample-resolved material for a node is the
% union of the sample-resolved intervals on its inbound edges. We then annotate
% the outbound edges with the intersection of their direct-inheritance intervals
% and the just-computed sample-resolved intervals for the node.
% For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
% carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
% joining \textsf{i} to its ancestor \textsf{k} then carries
% material on $(0, 2]$ since although \textsf{i} directly inherits interval
% $(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
% ancestral to the sample. For each node, we also keep track of the number of
% descendant samples over each genomic interval, and stop tracking that
% interval once the lineages from all samples have coalesced. For example,
% the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
% already fully coalesced (indicated by a fully black genomic region).

@@

% The process of converting a gARG with local-inheritance annotations
% to its sample-resolved equivalent is also known as ``simplification'',
% and is the key algorithm underlying recent
% advances in forwards-in-time
% simulation~\citep{kelleher2018efficient,haller2018tree}.
% It has many useful properties, as we discuss in the next section.
